\scuba: iterative map-making with the Sub-Millimetre User Reduction Facility

### 2012-09-12 smurf_paper.tex
Minor typos/corrections

Incoming light passes through a beam-splitter, and then bandpass
filters, providing simultaneous illumination and wavelength definition
of both the 450 and 850\,\micron\ focal planes. Each focal plan is
populated by four ``subarrays'' (labeled s4a--s4d at 450\,\micron, and
s8a--s8d at 850\,\micron), each consisting of 32 columns and 40 rows
of bolometers. The bolometers are thermal absorbers coupled to
superconducting transition-edge sensors (TESs) for thermometry.
Temperature variations in the TESs produce changing currents, and
therefore varying magnetic fields, which are detected and amplified
using chains of superconducting quantum interference devices (SQUIDs),
before the larger output currents are digitised. Each detector has its
own SQUID for the first-stage of the amplification, but the remaining
stages occur within a common chain of SQUIDs for each of the 32
columns. All 40 rows are read out in sequence at a row-visit frequency
of about 12\,kHz. Such a high sample rate is unnecessary to produce
maps, so the data are re-sampled to approximately 200\,Hz before
writing to disk. This rate provides a sample every 3.0\,arcsec, or
approximately one third of the 450\,\micron\ diffraction-limited
full-width at half-maximum (FWHM) -- a typical rule-of-thumb for
adequately sampling a Gaussian point spread function (PSF). There is
an additional 41st row of SQUID readouts that are not connected to
TESs. These ``dark SQUIDs'' track non-thermal noise sources that are
common to each column's amplifier chain. The relationship between the
output digitised current, $I$, and the input power, $P$, or $dI/dP$ is
established using flatfield observations immediately prior to science
observations, in which the output signal is measured throughout a ramp
of the pixel heaters (which provide a known input power); see
Section~2.1 in \citet{dempsey2012} and Section~5.3 in
\citet{holland2012} for more details. Finally, the conversion to
astronomical flux units from power involves a correction for
atmospheric extinction \citep[primarily using the JCMT Water Vapour
Monitor to track line-of-sight opacity variations, see Section~3
in][]{dempsey2012}, and the application of a flux conversion factor
(FCF) which is established from regular measurements of calibrators
such as Uranus \citep[Section~5 in][]{dempsey2012}.

@@

The approach taken by SMURF to reducing \scuba\ data is to model (and
remove) predominantly low-frequency noise sources that are correlated
amongst detectors, and to iterate this process along with estimates of
the map. Significant experimentation with different models for noise
sources during \scuba\ commissioning required a highly flexible
software framework, and a configurable user interface. To achieve this
goal, while minimizing development time, it was decided to build SMURF
as a Starlink package \citep{2009ASPC..411..418J}, which provides
access to a large suite of libraries (including a commanding and
messaging interface, astrometric coordinate conversions and generation
of standard WCS information, file formats etc.). Furthermore,
Starlink\footnote{\url{http://www.starlink.ac.uk}} is
open-source\footnote{\url{https://github.com/Starlink}}, and already
used extensively at the Joint Astronomy Centre (host of the JCMT) for
many other systems \citep{jenness2011}, which helps with
interoperability. Though originally written in FORTRAN, many of the
core Starlink libraries are now ported to native C, or at least have a
C interface. It was therefore decided to develop SMURF in C as well
(rather than C++, for example, which would have required adding
further dependencies to Starlink), although we have taken an
object-oriented approach. For example, all of the data for a given
noise model are encapsulated in a C structure, and there is a standard
interface for all functions that handle models (member data and
functions for the class, respectively). In this way it is easy to
extend SMURF with new models. Parallelisation is incorporated in the
most time-consuming low-level routines using threads (e.g., performing
Fast Fourier Transforms, re-gridding the data), usually handling
either independent blocks of bolometers, or blocks of time, in each
thread, depending on the nature of the calculation. We have found that
for typical data sets the processing time scales well with the number
of central processing unit (CPU) cores, although beyond 8 the returns
are diminished due to tasks that cannot be parallelised (e.g., reading
the data from disk).

@@

As an alternative, we also investigated an algorithm in which the FFT
of $x_i$ is simply truncated to the target sample rate before
transforming back to the time domain (i.e., applying a hard-edged
low-pass filter). In practice the noise performance was
indistinguishable, and required slightly more execution time using
this method. It is therefore not currently available to users.

@@

The accurate measurement of positions and brightnesses of known point
sources are necessary in real-time to establish telescope pointing
offsets and focus. They are also necessary to measure the FCF
(absolute calibration), and hence noise performance of the instrument
in astronomical units.  In this example we produce a 450\,\micron\ map
of Uranus (observation 26 on 20111017), which is a nearly point-like
source for \scuba\ that is commonly used as a primary flux
calibrator. The CV daisy pattern was used, with a scan speed of 155
arcsec\,sec$^{-1}$. We perform several different reductions of the
data to illustrate the purpose of various model components and the
convergence properties of the solution (Fig.~\ref{fig:pointmaps}). In
all cases the maps are produced on a grid of azimuth (horizontal) and
elevation (vertical) offsets from the position of Uranus (the origin),
using 2\,arcsec pixels.

@@

To suppress noise in the map, we construct a whitening filter from the
reciprocal of the jackknife angular power spectrum (orange line in
Fig.~\ref{fig:lockman_pspec}), normalised by the white-noise level
estimated from the RMS power at angular frequencies $>
0.1$\,arcsec$^{-1}$. The filter is applied by scaling the power
spectrum of the map by this function, and then transforming back into
real space. The whitened map is shown in Fig.~\ref{fig:lockman_maps}b,
and for comparison, the jackknife map has also been whitened in
Fig.~\ref{fig:lockman_maps}e. In both cases, the maps are visually
flatter than the non-whitened cases.

@@

Have real astronomical sources been detected using the matched filter?
For both the smoothed signal and jackknife maps, orange circles denote
3.8-$\sigma$ peaks. While not justified here, this threshold is fairly
typical for other ground-based submillimetre surveys in recent years
\citep[e.g.,][]{coppin2006,perera2008,2009ApJ...707.1201W} leading to
estimated false-identification rates of order $\sim$5\%, and is chosen
as a convenient reference. In the former, 10 peaks are found, whereas
there is only 1 in the latter. However, this test does not preclude
the possibility that some correlated noise made it into the jackknife
map, in which case the estimated \snr s are misleading.

@@

The transfer functions $G(k)$ are plotted as thin solid lines in the
right panel of Fig.~\ref{fig:m17_def_ps}. This formula produces
extremely noisy values at large frequencies (small scales), and we
therefore set it to a value of 1 above 0.015\,arcsec$^{-1}$, as well
as any point in the curve where $P_\mathrm{M}(k)$ exceeds
$P_\mathrm{S}(k)$ (i.e., $G(K)$ is assumed to be $\le$1). As expected,
the larger the scale of the filter, the lower the spatial frequency at
which the map transfer function falls. We then correct the map and
jackknife PSDs by dividing by $G(k)$ to produce the thick solid, and
dashed lines in the right panel of Fig.~\ref{fig:m17_def_ps}. This
shows us that, even though the raw noise in the left panel is lower at
small scales when a smaller-scale filter is used (e.g., the red dashed
line), once we correct for the transfer function, we actually win in a
\snr\ sense at large scales using the larger-scale filter (the
corrected noise is lower), as would be expected.

>>>>

Incoming light passes through a beam-splitter, and then bandpass
filters, providing simultaneous illumination and wavelength definition
of both the 450 and 850\,\micron\ focal planes. Each focal plan is
populated by four ``subarrays'' (labeled s4a--s4d at 450\,\micron, and
s8a--s8d at 850\,\micron), each consisting of 32 columns and 40 rows
of bolometers. The bolometers are thermal absorbers coupled to
superconducting transition-edge sensors (TESs) for thermometry.
Temperature variations in the TESs produce changing currents, and
therefore varying magnetic fields, which are detected and amplified
using chains of superconducting quantum interference devices (SQUIDs),
before the larger output currents are digitised. Each detector has its
own SQUID for the first-stage of the amplification, but the remaining
stages occur within a common chain of SQUIDs for each of the 32
columns. All 40 rows are read out in sequence at a row-visit frequency
of about 12\,kHz. Such a high sample rate is unnecessary to produce
maps, so the data are re-sampled to approximately 200\,Hz before
writing to disk. This rate provides a sample every 3.0\,arcsec, or
approximately one third of the 450\,\micron\ diffraction-limited
full-width at half-maximum (FWHM) -- a typical rule-of-thumb for
adequately sampling a Gaussian point spread function (PSF) -- at the maximum scanning speed. There is
an additional 41st row of SQUID readouts that are not connected to
TESs. These ``dark SQUIDs'' track non-thermal noise sources that are
common to each column's amplifier chain. The relationship between the
output digitised current, $I$, and the input power, $P$, or $dI/dP$ is
established using flatfield observations immediately prior to science
observations, in which the output signal is measured throughout a ramp
of the pixel heaters (which provide a known input power); see
Section~2.1 in \citet{dempsey2012} and Section~5.3 in
\citet{holland2012} for more details. Finally, the conversion to
astronomical flux units from power involves a correction for
atmospheric extinction \citep[primarily using the JCMT Water Vapour
Monitor to track line-of-sight opacity variations, see Section~3
in][]{dempsey2012}, and the application of a flux conversion factor
(FCF) which is established from regular measurements of calibrators
such as Uranus \citep[Section~5 in][]{dempsey2012}.

@@

The approach taken by SMURF to reducing \scuba\ data is to model (and
remove) predominantly low-frequency noise sources that are correlated
amongst detectors, and to iterate this process along with estimates of
the map. Significant experimentation with different models for noise
sources during \scuba\ commissioning required a highly flexible
software framework, and a configurable user interface. To achieve this
goal, while minimizing development time, it was decided to build SMURF
as a Starlink package \citep{2009ASPC..411..418J}, which provides
access to a large suite of libraries (including a commanding and
messaging interface, astrometric coordinate conversions and generation
of standard WCS information, file formats etc.). Furthermore,
Starlink\footnote{\url{http://www.starlink.ac.uk}} is
open-source\footnote{\url{https://github.com/Starlink}} (distributed under the GNU General Public Licence v3), and already
used extensively at the Joint Astronomy Centre (host of the JCMT) for
many other systems \citep{jenness2011}, which helps with
interoperability. Though originally written in FORTRAN, many of the
core Starlink libraries are now ported to native C, or at least have a
C interface. It was therefore decided to develop SMURF in C as well
(rather than C++, for example, which would have required adding
further dependencies to Starlink), although we have taken an
object-oriented approach. For example, all of the data for a given
noise model are encapsulated in a C structure, and there is a standard
interface for all functions that handle models (member data and
functions for the class, respectively). In this way it is easy to
extend SMURF with new models. Parallelisation is incorporated in the
most time-consuming low-level routines using threads (e.g., performing
Fast Fourier Transforms, re-gridding the data), usually handling
either independent blocks of bolometers, or blocks of time, in each
thread, depending on the nature of the calculation. We have found that
for typical data sets the processing time scales well with the number
of central processing unit (CPU) cores, although beyond 8 the returns
are diminished due to tasks that cannot be parallelised (e.g., reading
the data from disk).

@@

As an alternative, we also investigated an algorithm in which the FFT
of $x_i$ is simply truncated to the target sample rate before
transforming back to the time domain (i.e., applying a hard-edged
low-pass filter). In practice the noise performance was
indistinguishable, and required slightly longer execution time than using
the method above. Therefore this method is not used.

@@

The accurate measurement of positions and brightnesses of known point
sources are necessary in real-time to establish telescope pointing
offsets and focus. They are also necessary to measure the FCF
(absolute calibration), and hence noise performance of the instrument
in astronomically-useful units.  In this example we produce a 450\,\micron\ map
of Uranus (observation 26 on 20111017), which is a nearly point-like
source for \scuba\ that is commonly used as a primary flux
calibrator. The CV daisy pattern was used, with a scan speed of 155
arcsec\,sec$^{-1}$. We perform several different reductions of the
data to illustrate the purpose of various model components and the
convergence properties of the solution (Fig.~\ref{fig:pointmaps}). In
all cases the maps are produced on a grid of azimuth (horizontal) and
elevation (vertical) offsets from the position of Uranus (the origin),
using 2\,arcsec pixels.

@@

To suppress noise in the map, we construct a whitening filter from the
reciprocal of the jackknife angular power spectrum (orange line in
Fig.~\ref{fig:lockman_pspec}), normalised by the white-noise level
estimated from the RMS power at angular frequencies $>
0.1$\,arcsec$^{-1}$. The filter is applied by scaling the power
spectrum of the map by this function, and then transforming back into
real space. The whitened map is shown in Fig.~\ref{fig:lockman_maps}b,
and for comparison, the jackknife map has also been whitened in
Fig.~\ref{fig:lockman_maps}e. In both cases, the maps are visibly
flatter than the non-whitened cases.

@@

Have real astronomical sources been detected using the matched filter?
For both the smoothed signal and jackknife maps, blue circles denote
3.8-$\sigma$ peaks. While not justified here, this threshold is fairly
typical for other ground-based submillimetre surveys in recent years
\citep[e.g.,][]{coppin2006,perera2008,2009ApJ...707.1201W} leading to
estimated false-identification rates of order $\sim$5\%, and is chosen
as a convenient reference. In the former, 10 peaks are found, whereas
there is only 1 in the latter. However, this test does not preclude
the possibility that some correlated noise made it into the jackknife
map, in which case the estimated \snr s are misleading.

@@

The transfer functions $G(k)$ are plotted as thin solid lines in the
right panel of Fig.~\ref{fig:m17_def_ps}. This formula produces
extremely noisy values at large frequencies (small scales), and we
therefore set it to a value of 1 above 0.015\,arcsec$^{-1}$, as well
as any point in the curve where $P_\mathrm{M}(k)$ exceeds
$P_\mathrm{S}(k)$ (i.e., $G(k)$ is assumed to be $\le$1). As expected,
the larger the scale of the filter, the lower the spatial frequency at
which the map transfer function falls. We then correct the map and
jackknife PSDs by dividing by $G(k)$ to produce the thick solid, and
dashed lines in the right panel of Fig.~\ref{fig:m17_def_ps}. This
shows us that, even though the raw noise in the left panel is lower at
small scales when a smaller-scale filter is used (e.g., the red dashed
line), once we correct for the transfer function, we actually win in a
\snr\ sense at large scales using the larger-scale filter (the
corrected noise is lower), as would be expected.
